---
title: "20190716_VarianceInsteadOfEgenes"
author: "Ben Fair"
date: "7/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

For all the analyses I have using different levels of eGene character as a proxy for stabilizing selection (to get at what genes expression levels are important, and which are allowed to vary), I want to also do the same with coefficient of variation(CV) (as opposed to eGene character which requires that we can explain the variation by a cis-SNP). Even though CV does not require a heritable component to gene expression (which would therefore not be a target of stabilizing selection), it is more easy to measure CV than cis-eQTL mapping, and we know that 80% of gene expression heritability is trans anyway, so prioritizing stabilizing selection targets by presence of cis-eGene quality really only adds minimal evidence for heritability.


```{r load-libraries, message=F, warning=F}
library(tidyverse)
library(knitr)
library("edgeR")
library(stats)
library(corrplot)
library(gplots)
library("clusterProfiler")
library("org.Hs.eg.db")
library(enrichplot)




# Helper function reference in body of later function
rep.col<-function(x,n){
   matrix(rep(x,each=n), ncol=n, byrow=TRUE)
}

# Modified from the function in the PowerAnalysis Rmarkdown.
# Function to return RPKM table from chimp and human datasets (n=38 each)
# Use GenesToKeep argument to subset the same list that I compared for eQTLs
GetRPKMCountTable <-function(ChimpCountTableFile, HumanCountTableFile, SubsampleSize, GenesToKeep, ChimpSampleDrop=NULL, HumanSampleDrop=NULL)
  #if SubsampleSize parameter == 0, use full table, otherwise, subsample from it
  {
   FullChimpData <- read.table(gzfile(ChimpCountTableFile), header=T, check.names=FALSE, skip=1)
   FullHumanData <- read.table(gzfile(HumanCountTableFile), header=T, check.names=FALSE, skip=1)
   
  if (!is.null(ChimpSampleDrop)){
     FullChimpData <- FullChimpData %>% dplyr::select(-ChimpSampleDrop)
   }   
  if (!is.null(HumanSampleDrop)){
     FullHumanData <- FullHumanData %>% dplyr::select(-HumanSampleDrop)
   }

   if (SubsampleSize==0){
     CountTableChimp <- FullChimpData
     colnames(CountTableChimp) <- paste0("C.", colnames(CountTableChimp))
     CountTableHuman <- FullHumanData
     colnames(CountTableHuman) <- paste0("H.", colnames(CountTableHuman))
     
   } else {
     CountTableChimp <- FullChimpData %>% dplyr::select(c(1:6, sample(7:length(FullChimpData), SubsampleSize)))
     colnames(CountTableChimp) <- paste0("C.", colnames(CountTableChimp))
     
     CountTableHuman <- FullHumanData %>% dplyr::select(c(1:6, sample(7:length(FullHumanData), SubsampleSize)))
     colnames(CountTableHuman) <- paste0("H.", colnames(CountTableHuman))
   }
   
CombinedTable <- inner_join(CountTableChimp[,c(1,7:length(CountTableChimp))], CountTableHuman[,c(1,7:length(CountTableHuman))], by=c("C.Geneid"="H.Geneid")) %>%
  column_to_rownames("C.Geneid") %>% as.matrix()

SpeciesFactor <- colnames(CombinedTable) %>% substr(1,1) %>% factor() %>% unclass() %>% as.character()

d0 <- DGEList(CombinedTable)
d0 <- calcNormFactors(d0)



d <- d0[GenesToKeep,]
mm <- model.matrix(~0 + SpeciesFactor)
# y <- voom(d, mm, normalize.method="cyclicloess", plot=F)

# GeneLengths <- inner_join(CountTableChimp[,c("C.Geneid", "C.Length")], CountTableHuman[,c("H.Geneid", "H.Length")], by=c("C.Geneid"="H.Geneid"))
# GeneLengthMatrix <- cbind(
#   rep.col(log2(GeneLengths$C.Length/1000), length(CountTableChimp)-6),
#   rep.col(log2(GeneLengths$H.Length/1000), length(CountTableHuman)-6))
# rownames(GeneLengthMatrix) <- GeneLengths$C.Geneid
# y$E <- y$E - GeneLengthMatrix[rownames(y$E),]

return(cpm(d, log=T))
# return(d0)

}

```


Get list of cis-eQTL tested genes that were tested in both chimps and humans and one-to-one orthologs and also in DE gene count tables (from reads mapped to ortho exons):

```{r}

CountTableChimpFile <- '../output/PowerAnalysisFullCountTable.Chimp.subread.txt.gz'
CountTableHumanFile <- '../output/PowerAnalysisFullCountTable.Human.subread.txt.gz'

eQTLs <- read.table(gzfile("../data/PastAnalysesDataToKeep/20190521_eQTLs_250kB_10MAF.txt.gz"), header=T)

# List of chimp tested genes
ChimpTestedGenes <- rownames(read.table('../output/ExpressionMatrix.un-normalized.txt.gz', header=T, check.names=FALSE, row.names = 1))

ChimpToHumanGeneMap <- read.table("../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", header=T, sep='\t', stringsAsFactors = F)

# Of this ortholog list, how many genes are one2one
OneToOneMap <- ChimpToHumanGeneMap %>%
  filter(Chimpanzee.homology.type=="ortholog_one2one")

# Read gtex heart egene list
# Only consider those that were tested in both species and are one2one orthologs
GtexHeartEgenes <- read.table("../data/Heart_Left_Ventricle.v7.egenes.txt.gz", header=T, sep='\t', stringsAsFactors = F) %>%
  mutate(gene_id_stable = gsub(".\\d+$","",gene_id)) %>%
  filter(gene_id_stable %in% OneToOneMap$Gene.stable.ID) %>%
  mutate(chimp_id = plyr::mapvalues(gene_id_stable, OneToOneMap$Gene.stable.ID,  OneToOneMap$Chimpanzee.gene.stable.ID, warn_missing = F)) %>%
  filter(chimp_id %in% ChimpTestedGenes)


EgenesTested <- gsub("\\..+", "", GtexHeartEgenes$gene_id, perl=T)
length(EgenesTested)
GenesInDESet <- read.table(gzfile(CountTableChimpFile), header=T, check.names=FALSE, skip=1)$Geneid
length(GenesInDESet)

GeneList <- intersect(as.character(GenesInDESet),EgenesTested)
kable(head(GeneList))
length(GeneList)
```

Ok, now get CountTable of 38 chimps, 38 humans (based on remapped data that maps to orthologous exons) of RPKM, and filtered for the outlier samples that I also left out of the DE gene power analysis

```{r}
HumanSamplesToDrop <- c(c("SRR1507229","SRR603918", "SRR1478149", "SRR598509", "SRR613186"), c("SRR1489693", "SRR598148", "59167", "SRR1478900", "SRR1474730", "61317"))
ChimpSamplesToDrop <- c("Little_R")




CountTable <- GetRPKMCountTable(CountTableChimpFile,
              CountTableHumanFile,
              0, GeneList, ChimpSampleDrop=ChimpSamplesToDrop, HumanSampleDrop = HumanSamplesToDrop)


qplot(apply(CountTable,1,mean), sqrt(apply(CountTable,1,var)), alpha=0.05) +
  geom_smooth(method="loess", show_guide = FALSE, se=F) +
  xlab("Mean expression (log(CPM)") +
  ylab("Standard deviation") +
  theme_bw() +
  theme(legend.position = "none")

scatter.smooth(apply(CountTable,1,mean), sqrt(apply(CountTable,1,var)), color="red")

CountTableNonLog <- 2**CountTable


# qplot(apply(CountTableNonLog,1,mean), sqrt(apply(CountTableNonLog,1,var)/apply(CountTableNonLog,1,mean)), alpha=0.01)


# ToPlot <- data.frame(CV=sqrt(apply(CountTableNonLog,1,var)/apply(CountTableNonLog,1,mean)),
#                      mean = apply(CountTableNonLog,1,mean), 
#                      SDlogexpression = sqrt(apply(CountTable,1,var)),
#                      logmean = apply(CountTable,1,mean))
# 
# # standard deviation vs expression (after log transforming RPKM), with CV as colors
# ggplot(ToPlot, aes(x=log(mean), y=SDlogexpression, color=log(CV))) +
#   geom_point()


```

Coefficient of variation still doesn't really capture the character I am going after (since all the high CV genes are highly expressed), which is how much does a gene vary relative to other genes at its expression level. For now (for simplicity) I may simply use the standard deviation (after log transformation as my metric to get at stabilizing selection)

Below I will try fitting a loess mean-variance trend and rank genes according the how elevated they are above that trend

```{r}

Gene.summarystats <- data.frame(CV=sqrt(apply(CountTableNonLog,1,var)/apply(CountTableNonLog,1,mean)),
                     mean = apply(CountTableNonLog,1,mean), 
                     SDlogexpression = sqrt(apply(CountTable,1,var)),
                     logmean = apply(CountTable,1,mean))

Gene.summarystats.lo <- loess(SDlogexpression ~ logmean, Gene.summarystats)
Gene.summarystats$SD.minus.loess <- Gene.summarystats$SDlogexpression - predict(Gene.summarystats.lo, Gene.summarystats$logmean)
ggplot(Gene.summarystats, aes(x=logmean, y=SDlogexpression, color=SD.minus.loess)) +
  geom_point() +
  geom_smooth(method="loess", show_guide = FALSE, se=F) +
  xlab("Mean expression (log(CPM)") +
  ylab("Standard deviation") +
  scale_colour_gradient2(name = "Residual from loess curve") +
  theme_bw() +
  theme(legend.position="bottom")


```

Alright that worked well, now do it for chimp, and again for human, and compare...

```{r}
Get.GeneSummaryStat.df <- function(Input.df){
  My.Gene.summarystats <- data.frame(mean = apply(Input.df,1,mean),
                                  SD = sqrt(apply(Input.df,1,var)))
  My.Gene.summarystats.lo <- loess(SD ~ mean, My.Gene.summarystats)
  My.Gene.summarystats$SD.minus.loess <- My.Gene.summarystats$SD - predict(My.Gene.summarystats.lo, My.Gene.summarystats$mean)
  return(My.Gene.summarystats)
}


CountTable.chimp <- CountTable %>% as.data.frame() %>% dplyr::select(contains("C."))
Chimp.summarystats <- Get.GeneSummaryStat.df(CountTable.chimp)

CountTable.human <- CountTable %>% as.data.frame() %>% dplyr::select(contains("H."))
Human.summarystats <- Get.GeneSummaryStat.df(CountTable.human)

# How does this statistic (std dev minus loess prediction) correlate across species
R<-cor(Chimp.summarystats$SD.minus.loess, Human.summarystats$SD.minus.loess, method="pearson")
lb1 <- paste("~R^2==~", round(R**2,2))
qplot(Chimp.summarystats$SD.minus.loess, Human.summarystats$SD.minus.loess, alpha=0.05) +
  xlab("Chimp") +
  ylab("Human")+
  annotate("text",x=Inf,y=-Inf, label=lb1, hjust=1, vjust=-1, parse=TRUE) +
  theme_bw() +
  theme(legend.position = "none")

#Plot of this statistic vs expression level:
R<-cor(Chimp.summarystats$mean, Chimp.summarystats$SD.minus.loess, method="pearson")
lb1 <- paste("~R^2==~", round(R**2,2))
ggplot(Chimp.summarystats, aes(x=mean, y=SD.minus.loess)) +
  geom_point(alpha=0.05) +
  annotate("text",x=Inf,y=-Inf, label=lb1, hjust=1, vjust=-1, parse=TRUE) +
  theme_bw()

```

That correlation is surprisingly good. I should check that it isn't a bug due to sample labels getting switched. Lets make correlation matrix of the original table to ensure that the chimp samples separate from the human samples.

```{r}
cor(CountTable$E, method = c("spearman")) %>%
heatmap.2(trace="none")
```

Ok wow, I guess it really is the case that gene expression variance is very consistent. If there really isn't a bug, and this statistic is measuring what it should, I expect it to correlate strongly to other measures of conservation like dN/dS or percent identity

```{r}
ChimpToHumanGeneMap <- read.table("../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", header=T, sep='\t', stringsAsFactors = F) %>% distinct(Gene.stable.ID, .keep_all = T)


ToPlot <- Chimp.summarystats %>%
  rownames_to_column() %>%
  left_join(ChimpToHumanGeneMap, by=c("rowname" = "Gene.stable.ID")) %>%
  mutate(rank = dense_rank(dplyr::desc(SD.minus.loess))) %>% 
  mutate(group = case_when(
                           rank>=500 ~ "NotTop500VarianceGene",
                           rank<=500 ~ "Top500VarianceGene")) %>%
  mutate(dN.dS = dN.with.Chimpanzee/dS.with.Chimpanzee)

ggplot(ToPlot, aes(color=group, x=100-X.id..query.gene.identical.to.target.Chimpanzee.gene+0.001)) +
  stat_ecdf(geom = "step") +
  scale_x_continuous(trans='log1p', limits=c(0,100), expand=expand_scale()) +
  ylab("Cumulative frequency") +
  xlab("Percent non-identitical amino acid between chimp and human") +
  annotate("text", x = 10, y = 0.4, label = paste("Mann-Whitney\none-sided P =", signif(wilcox.test(data=ToPlot, X.id..query.gene.identical.to.target.Chimpanzee.gene ~ group, alternative="greater")$p.value, 2) )) +
  theme_bw() +
  theme(legend.position = c(.80, .2), legend.title=element_blank())

#What is overall correlation (previous analysis, where I compared two groups based on a threshold categorization of the other variable, is not the most sensitive method)
cor.test(ToPlot$X.id..query.gene.identical.to.target.Chimpanzee.gene, ToPlot$SD.minus.loess, method='spearman')
ggplot(ToPlot, aes(y=SD.minus.loess, x=100-X.id..query.gene.identical.to.target.Chimpanzee.gene+0.001)) +
  geom_point() +
  scale_x_continuous(trans='log1p', limits=c(0,100), expand=expand_scale()) +
  ylab("SD.minus.loess") +
  xlab("Percent non-identitical amino acid between chimp and human") +
  theme_bw() +
  theme(legend.position = c(.80, .2), legend.title=element_blank())


ggplot(ToPlot, aes(color=group,x=dN.dS)) +
  stat_ecdf(geom = "step") +
  ylab("Cumulative frequency") +
  xlab("dN/dS") +
  scale_x_continuous(trans='log10', limits=c(0.01,10)) +
  annotate("text", x = 1, y = 0.4, label = paste("Mann-Whitney\none-sided P =", signif(wilcox.test(data=ToPlot, dN.dS ~ group, alternative="less")$p.value, 2) )) +
  theme_bw() +
  theme(legend.position = c(.80, .2), legend.title=element_blank())

```

Ok now subtract the variance metric (Chimp - human) to get a ordered list of genes where higher numbers means more variance in chimp.

```{r}
RankedGeneList<-Chimp.summarystats$SD.minus.loess - Human.summarystats$SD.minus.loess
names(RankedGeneList) <- rownames(Chimp.summarystats)
SortedGeneList <- sort(RankedGeneList, decreasing=T)

```

GSEA analysis


```{r}

#bp  
gsego.cc.chimp.high.var <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "BP",
                 nPerm = 100000)
A<-as.data.frame(gsego.cc.chimp.high.var)
bp2 <- simplify(gsego.cc.chimp.high.var, cutoff=0.7, by="p.adjust", select_fun=min)

gseaplot2(gsego.cc.chimp.high.var, geneSetID = c("GO:0002263", "GO:0061061", "GO:0072538", "GO:0000076"), title="ChimpVariance - HumanVariance", pvalue_table = TRUE)

dotplot(bp2, font.size=8, showCategory=10)


#all
gsego.all.chimp.high.var <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "ALL",
                 nPerm = 100000)
A<-as.data.frame(gsego.all.chimp.high.var)
A %>% as.data.frame() %>% dplyr::select(ONTOLOGY, ID, Description, enrichmentScore) %>% arrange(desc(abs(enrichmentScore))) %>% head(20)

A %>% as.data.frame() %>% dplyr::select(ONTOLOGY, ID, Description, enrichmentScore) %>% arrange(desc(abs(enrichmentScore))) %>% head(20)

# bp2 <- simplify(gsego.all.chimp.high.var, cutoff=0.7, by="p.adjust", select_fun=min)

# bp2 %>% as.data.frame() %>% dplyr::select(ONTOLOGY, ID, Description, enrichmentScore) %>% arrange(desc(abs(enrichmentScore))) %>% head(20)



table(A$ONTOLOGY)
A$ID

```

Keep in mind GSEA finds both enrichment at both top and bottom of the list. MHC complex is on here, with higher enrichment in chimp

```{r}
CC.gsea <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "CC",
                 nPerm = 100000)

BP.gsea <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "BP",
                 nPerm = 1000000)

MF.gsea <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "MF",
                 nPerm = 100000)

CC.gsea.simplified <- as.data.frame(simplify(CC.gsea))
CC.gsea.simplified$OntologyCategory <- "Cellular.Component"
BP.gsea.simplified <- as.data.frame(simplify(BP.gsea))
BP.gsea.simplified$OntologyCategory <- "Biological.Process"
MF.gsea.simplified <- as.data.frame(simplify(MF.gsea))
MF.gsea.simplified$OntologyCategory <- "Molecular.Function"

Combined <- rbind(
      CC.gsea.simplified,
      BP.gsea.simplified,
      MF.gsea.simplified)

Combined %>% 
  group_by(OntologyCategory) %>%
  top_n(n = 5, wt = abs(enrichmentScore)) %>%
  ungroup() %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(OntologyCategory~., scales = "free") +
  scale_colour_gradient(low="red", high="black") +
  facet_grid(OntologyCategory~., scales = "free") +
  labs(color = "Adjusted P-value") +
  theme_bw()

Combined %>% 
  group_by(OntologyCategory) %>%
  top_n(n = -7, wt = qvalues) %>%
  top_n(n = 7, wt = setSize) %>%
  ungroup() %>%
  # group_by(OntologyCategory) %>%
  # sample_n(8) %>%
  ggplot(aes(x=enrichmentScore, y=Description, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(OntologyCategory~., scales = "free") +
  # scale_colour_gradient(low="red", high="black") +
  # labs(color = "Adjusted P-value") +
  theme_bw()


```

Also worth asking what gene categories are at top of list in both species... Since the mean-adjusted-variance (residual from loess) metric seems to correlate so well. It might help to interpret what that means if I knew of the significant gene categories... How about we create an ordered gene list based on that average of this residual-from-loess metric between chimp and human...

```{r}
RankedGeneList<-rowMeans(cbind(Chimp.summarystats$SD.minus.loess,Human.summarystats$SD.minus.loess))
names(RankedGeneList) <- rownames(Chimp.summarystats)
SortedGeneList <- sort(RankedGeneList, decreasing=T)



BP.gsea <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "BP",
                 nPerm = 1000000)

MF.gsea <- gseGO(gene = SortedGeneList,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 maxGSSize = 500,
                 ont = "MF",
                 nPerm = 100000)

CC.gsea.simplified <- as.data.frame(simplify(CC.gsea))
CC.gsea.simplified$OntologyCategory <- "Cellular.Component"
BP.gsea.simplified <- as.data.frame(simplify(BP.gsea))
BP.gsea.simplified$OntologyCategory <- "Biological.Process"
MF.gsea.simplified <- as.data.frame(simplify(MF.gsea))
MF.gsea.simplified$OntologyCategory <- "Molecular.Function"

Combined <- rbind(
      CC.gsea.simplified,
      BP.gsea.simplified,
      MF.gsea.simplified)

Combined %>% 
  group_by(OntologyCategory) %>%
  top_n(n = 5, wt = abs(enrichmentScore)) %>%
  ungroup() %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(OntologyCategory~., scales = "free") +
  scale_colour_gradient(low="red", high="black") +
  facet_grid(OntologyCategory~., scales = "free") +
  labs(color = "Adjusted P-value") +
  theme_bw()

Combined %>% 
  group_by(OntologyCategory) %>%
  top_n(n = -7, wt = qvalues) %>%
  top_n(n = 7, wt = setSize) %>%
  ungroup() %>%
  # group_by(OntologyCategory) %>%
  # sample_n(8) %>%
  ggplot(aes(x=enrichmentScore, y=Description, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(OntologyCategory~., scales = "free") +
  # scale_colour_gradient(low="red", high="black") +
  # labs(color = "Adjusted P-value") +
  theme_bw()


```



In addition to GSEA analysis, it might be worthwile to also try to correlate difference in variance metric to chromatin-interaction score:

```{r}
SampleA<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_A-21792_10kb_norm.gz"), sep='\t')
SampleB<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_B-28126_10kb_norm.gz"), sep='\t')
SampleC<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_C-3649_10kb_norm.gz"), sep='\t')
SampleD<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_D-40300_10kb_norm.gz"), sep='\t')
SampleE<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_E-28815_10kb_norm.gz"), sep='\t')
SampleF<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_F-28834_10kb_norm.gz"), sep='\t')
SampleG<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_G-3624_10kb_norm.gz"), sep='\t')
SampleH<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_H-3651_10kb_norm.gz"), sep='\t')


ChimpToHumanGeneMap <- read.table("../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", header=T, sep='\t', stringsAsFactors = F)
kable(head(ChimpToHumanGeneMap))

# Of this ortholog list, how many genes are one2one
table(ChimpToHumanGeneMap$Chimpanzee.homology.type)
OneToOneMap <- ChimpToHumanGeneMap %>%
  filter(Chimpanzee.homology.type=="ortholog_one2one")
ChimpToHuman.ID <- function(Chimp.ID){
  #function to convert chimp ensembl to human ensembl gene ids
  return(
    plyr::mapvalues(Chimp.ID, OneToOneMap$Chimpanzee.gene.stable.ID, OneToOneMap$Gene.stable.ID, warn_missing = F)
  )}

HumanInteractions <- data.frame(H.Score = base::rowSums(cbind(SampleA, SampleB, SampleE, SampleF))) %>%
  rownames_to_column() %>%
  mutate(HumanID = gsub("(.+?)\\..+?", "\\1", rowname, perl=T))


ChimpInteractions <- data.frame(C.Score = rowSums(cbind(SampleC, SampleD, SampleG, SampleH))) %>%
  rownames_to_column() %>%
  mutate(HumanID = ChimpToHuman.ID(rowname))


ToPlot <- data.frame(SpeciesVarianceDiff = RankedGeneList) %>%
  rownames_to_column() %>%
  left_join(HumanInteractions, by=c("rowname"="HumanID")) %>%
  left_join(ChimpInteractions, by=c("rowname"="HumanID")) %>% 
  mutate(InteractionDifference=H.Score - C.Score) %>%
  filter(!is.na(H.Score)) %>%
  filter(!is.na(C.Score))

ggplot(ToPlot, aes(x=SpeciesVarianceDiff, y=InteractionDifference)) +
  geom_point() +
  theme_bw() +
  xlab("Variation in expression\nTighter in human <--  --> Tighter in chimp") +
  ylab("Differential contacts in cis window\nMore in human <--  --> More in chimp") +
  geom_smooth(method='lm',formula=y~x)

cor.test(x=ToPlot$SpeciesVarianceDiff, y=ToPlot$InteractionDifference, method="spearman")

contacts.v.eGene.lm = lm(InteractionDifference ~ SpeciesVarianceDiff, data=ToPlot)
summary(contacts.v.eGene.lm)
plot(contacts.v.eGene.lm)

```

Not significant. Perhaps most intuitive explanation is to why this relationship was significant with eGenes (species difference in neighborhood chromatin contacts partly explains species difference in cis-eGene rank) but not for variance (species difference in neghborhood chromatin contacts does not significantly explain any difference in within- species variance) is that the chromatin contacts only mediate cis-variance, while 80% of expression variance is in trans.
