---
title: "20190627_DiffContactsWithEgenes"
author: "Ben Fair"
date: "6/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r load-libraries}
library(tidyverse)
library(knitr)
library("edgeR")
library(corrplot)
library(gplots)
library(pROC)
library(qvalue)
library(reshape2)
library(gridExtra)
```

As an initial pass at testing the hypothesis that Chimp-specific cis-eGenes have more DNA contacts at those cis-windows than humans in those same cis-windows, I looked at chimp/human differential contact windows from Eres et al and intersected in with my list of Chimp eGene locations (+/- 100kb). Then I looked at the sum of effect sizes within each of those windows.

```{r}
DataIn <- read.table("../data/DCContactsInEgenes.bed", col.names = c("chr", "start", "stop", "gene", "blank", "strand", "chrContact", "startContact", "beta", "nameContact", "beta", "strandContact"), sep='\t', row.names = NULL)

GroupedData <- DataIn %>%
  group_by(stop) %>%
  summarise(BetaSum = sum(beta))

ggplot(GroupedData, aes(x=BetaSum)) +
  stat_ecdf(geom = "step") +
  xlab("Differential contact effect size sum over Chimp eGenes cis-window\n(Negative means more contact in chimp)") +
  ylab("Cumulative frequency") +
  theme_bw()

wilcox.test(GroupedData$BetaSum, mu = 0, alternative = "less")
```


Ok it seems there is a subtle but significant shift in that chimp eGenes have more differential contacts a way that slightly favors high connectivity in chimps.

To explore this more carefully, Ittai gave me a list of Homer-normalized contact data (not just significant differential contacts and effect sizes) for all his 8 samples (4 human, 4 chimp) for the cis-window surrounding each eGene (+/- 250kb).

From this I can estimate connectivity within a cis window as the sum of all Homer normalized contact scores within a species. Then I will look at the difference between that sum between species, and ask if it is correlated with some species measure of eGene character.

First, read in Ittai's Homer normalized contact data for each individual in both chimp and human...

```{r}
SampleA<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_A-21792_10kb_norm.gz"), sep='\t')
SampleB<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_B-28126_10kb_norm.gz"), sep='\t')
SampleC<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_C-3649_10kb_norm.gz"), sep='\t')
SampleD<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_D-40300_10kb_norm.gz"), sep='\t')
SampleE<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_E-28815_10kb_norm.gz"), sep='\t')
SampleF<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_F-28834_10kb_norm.gz"), sep='\t')
SampleG<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_G-3624_10kb_norm.gz"), sep='\t')
SampleH<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_H-3651_10kb_norm.gz"), sep='\t')

HumanInteractions <- data.frame(H.Score = rowSums(cbind(SampleA, SampleB, SampleE, SampleF))) %>%
  rownames_to_column() %>%
  mutate(HumanID = gsub("(.+?)\\..+?", "\\1", rowname, perl=T))


ChimpInteractions <- data.frame(C.Score = rowSums(cbind(SampleC, SampleD, SampleG, SampleH))) %>%
  rownames_to_column("ChimpID")


```


Ok now read in eQTL data...

```{r}
eQTLs <- read.table(gzfile("../data/PastAnalysesDataToKeep/20190521_eQTLs_250kB_10MAF.txt.gz"), header=T)

# List of chimp tested genes
ChimpTestedGenes <- rownames(read.table('../output/ExpressionMatrix.un-normalized.txt.gz', header=T, check.names=FALSE, row.names = 1))

ChimpToHumanGeneMap <- read.table("../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", header=T, sep='\t', stringsAsFactors = F)
kable(head(ChimpToHumanGeneMap))

# Of this ortholog list, how many genes are one2one
table(ChimpToHumanGeneMap$Chimpanzee.homology.type)
OneToOneMap <- ChimpToHumanGeneMap %>%
  filter(Chimpanzee.homology.type=="ortholog_one2one")

# Read gtex heart egene list
# Only consider those that were tested in both species and are one2one orthologs
GtexHeartEgenes <- read.table("../data/Heart_Left_Ventricle.v7.egenes.txt.gz", header=T, sep='\t', stringsAsFactors = F) %>%
  mutate(gene_id_stable = gsub(".\\d+$","",gene_id)) %>%
  filter(gene_id_stable %in% OneToOneMap$Gene.stable.ID) %>%
  mutate(chimp_id = plyr::mapvalues(gene_id_stable, OneToOneMap$Gene.stable.ID,  OneToOneMap$Chimpanzee.gene.stable.ID, warn_missing = F)) %>%
  filter(chimp_id %in% ChimpTestedGenes)

ChimpToHuman.ID <- function(Chimp.ID){
  #function to convert chimp ensembl to human ensembl gene ids
  return(
    plyr::mapvalues(Chimp.ID, OneToOneMap$Chimpanzee.gene.stable.ID, OneToOneMap$Gene.stable.ID, warn_missing = F)
  )}
```


First question: do the ~300 chimp eGenes have more contacts in their cis-window in chimp

```{r}

Chimp_OrderedGenes <- eQTLs %>%
    group_by(gene) %>% 
    dplyr::slice(which.min(qvalue)) %>%
    filter(gene %in% GtexHeartEgenes$chimp_id) %>%
    left_join(GtexHeartEgenes, by=c("gene"="chimp_id")) %>%
    dplyr::select(gene, qvalue, qval) %>% as.data.frame() %>%
    mutate(ChimpRank = dense_rank(qvalue)) %>%
    mutate(HumanRank = dense_rank(qval)) %>%
    mutate(RankDifference = HumanRank-ChimpRank) %>%
    filter(qvalue <0.1) %>%
    mutate(HumanID=ChimpToHuman.ID(gene))

# OneToOneMap %>%
#   inner_join(HumanInteractions, by=c("Gene.stable.ID"="HumanId")) %>% dim()
#   inner_join(ChimpInteractions, by=c("Chimpanzee.gene.stable.ID"="ChimpID")) %>% dim()
#   right_join(Chimp_OrderedGenes, by=c("Chimpanzee.gene.stable.ID"="gene")) %>% dim()

Chimp_OrderedGenes.WithContactInfo <- Chimp_OrderedGenes %>%
  left_join(HumanInteractions, by=c("HumanID")) %>%
  left_join(ChimpInteractions, by=c("gene"="ChimpID")) %>% 
  mutate(InteractionDifference=H.Score - C.Score)

  
ggplot(Chimp_OrderedGenes.WithContactInfo, aes(x=InteractionDifference)) +
  stat_ecdf(geom = "step") +
  xlab("Difference in contacts over chimp eGene cis-windows\n(Positive means more contact in chimp)") +
  ylab("Cumulative frequency") +
  theme_bw()
  
ggplot(Chimp_OrderedGenes.WithContactInfo, aes(x=RankDifference, y=InteractionDifference)) +
  geom_point() +
  theme_bw() +
  xlab("Rank Difference in eGene significance\nMore in human <--  --> More in chimp") +
  ylab("Differential contacts in cis window\nMore in human <--  --> More in chimp") +
  geom_smooth(method='lm',formula=y~x)

cor.test(x=Chimp_OrderedGenes.WithContactInfo$RankDifference, y=Chimp_OrderedGenes.WithContactInfo$InteractionDifference, method="spearman")

contacts.v.eGene.lm = lm(InteractionDifference ~ RankDifference, data=Chimp_OrderedGenes.WithContactInfo)
summary(contacts.v.eGene.lm)
plot(contacts.v.eGene.lm)
```

Ok yes, there is a slight correlation between chimp eGene character, and contacts in chimp. Now I am going to take a different approach to ask a similar question: Do lead chimp eQTLs have more contacts between a gene and its promoter window than human.

