---
title: "DE genes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

## Overview:

Do DE gene analysis for chimp/human. Leave out virus challanged chimps


## Analysis

Load necessary libraries:

```{r load-libraries, message=F, warning=F}
library(tidyverse)
library(knitr)
library("edgeR")
library(corrplot)
library(gplots)
library(pROC)
library(qvalue)
library(reshape2)
library(gridExtra)
library(readxl)

source("../code/CustomFunctions.R")
```


Read in count tables

```{r read-data}
CountTableChimpFile <- '../output/PowerAnalysisFullCountTable.Chimp.subread.txt.gz'
CountTableHumanFile <- '../output/PowerAnalysisFullCountTable.Human.subread.txt.gz'
```

Drop samples that are to be dropped for various reasons. see [power analysis](20190613_PowerAnalysis.html). Also drop the virus challanged chimps

```{r}
HumanSamplesToDrop <- c(c("SRR1507229","SRR603918", "SRR1478149", "SRR598509", "SRR613186"), c("SRR1489693", "SRR598148", "59167", "SRR1478900", "SRR1474730", "61317"))
OtherMetadata <- as.data.frame(read_excel("../data/Metadata.xlsx"))
VirusChallengedChimps <- OtherMetadata %>% filter(grepl("V+",Viral.status)) %>% pull(Individual.ID)
ChimpSamplesToDrop <- c(c("Little_R"), VirusChallengedChimps)
```

See [source code](https://github.com/bfairkun/Comparative_eQTL/blob/master/code/CustomFunctions.R) for DE gene analysis or refer to basically the same code in my [power analysis](20190613_PowerAnalysis.html).

```{r}
CountTableChimp <- read.table(gzfile(CountTableChimpFile), header=T, check.names=FALSE, skip=1) %>% dplyr::select(-ChimpSamplesToDrop)
CountTableHuman <- read.table(gzfile(CountTableHumanFile), header=T, check.names=FALSE, skip=1) %>% dplyr::select(-HumanSamplesToDrop)
CountTableFull <- inner_join(CountTableChimp[,c(1,7:length(CountTableChimp))], CountTableHuman[,c(1,7:length(CountTableHuman))], by="Geneid") %>%
  column_to_rownames("Geneid") %>% as.matrix() %>% DGEList() %>% calcNormFactors()
cutoff <- 6
drop <- which(apply(cpm(CountTableFull), 1, max) < cutoff)
d <- CountTableFull[-drop,] 

#Interval for null hypothesis
FC.NullInterval <- log2(1)

#True results are those using all samples
True.efit <- DE.Subsampled(CountTableChimpFile,
              CountTableHumanFile,
              0, FC.NullInterval, drop, ChimpSampleDrop=ChimpSamplesToDrop, HumanSampleDrop = HumanSamplesToDrop)
TrueResponse <- decideTests(True.efit, p.value=0.05)

WriteOut <- data.frame(coefficients=as.numeric(True.efit$coefficients), pval=as.numeric(True.efit$p.value), signif=decideTests(True.efit))
ggplot(WriteOut, aes(x=coefficients, y=-log10(pval), color=factor(DE))) +
  geom_point(alpha=0.1) +
  scale_x_continuous(limits=c(-5,5))

```

Write out results to data

```{r, eval=F}
WriteOut %>% tibble::rownames_to_column("gene") %>%
write.table("../data/DE_genes.NoVirusChallangedInds.txt", sep='\t', row.names = F, col.names = T, quote=F)
```


