---
title: "20190627_DiffContactsWithEgenes"
author: "Ben Fair"
date: "6/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r load-libraries}
library(tidyverse)
library(knitr)
library(corrplot)
library(gplots)
library(pROC)
library(qvalue)
library(reshape2)
library(gridExtra)
```




```{r}
SampleA<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_A-21792_10kb_norm.gz"), sep='\t')
SampleB<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_B-28126_10kb_norm.gz"), sep='\t')
SampleC<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_C-3649_10kb_norm.gz"), sep='\t')
SampleD<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_D-40300_10kb_norm.gz"), sep='\t')
SampleE<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_E-28815_10kb_norm.gz"), sep='\t')
SampleF<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_F-28834_10kb_norm.gz"), sep='\t')
SampleG<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_G-3624_10kb_norm.gz"), sep='\t')
SampleH<- read.csv(gzfile("../data/IttaiHomerInteractionScoresInCisWindows/adj_bins_25_H-3651_10kb_norm.gz"), sep='\t')

HumanInteractions <- data.frame(H.Score = rowSums(cbind(SampleA, SampleB, SampleE, SampleF))) %>%
  rownames_to_column() %>%
  mutate(HumanID = gsub("(.+?)\\..+?", "\\1", rowname, perl=T))


ChimpInteractions <- data.frame(C.Score = rowSums(cbind(SampleC, SampleD, SampleG, SampleH))) %>%
  rownames_to_column("ChimpID")


```


Ok now read in eQTL data...

```{r}
eQTLs <- read.table(gzfile("../data/PastAnalysesDataToKeep/20190521_eQTLs_250kB_10MAF.txt.gz"), header=T)

# List of chimp tested genes
ChimpTestedGenes <- rownames(read.table('../output/ExpressionMatrix.un-normalized.txt.gz', header=T, check.names=FALSE, row.names = 1))

ChimpToHumanGeneMap <- read.table("../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", header=T, sep='\t', stringsAsFactors = F)
kable(head(ChimpToHumanGeneMap))

# Of this ortholog list, how many genes are one2one
table(ChimpToHumanGeneMap$Chimpanzee.homology.type)
OneToOneMap <- ChimpToHumanGeneMap %>%
  filter(Chimpanzee.homology.type=="ortholog_one2one")

HumanLeadSnps <- read.table(gzfile('../data/Heart_Left_Ventricle.v7.250kB.leadsnps.txt.gz'), col.names = c("gene", "snp", "tss.dist", "ma_samples", "ma_count", "maf", "p", "slope", "slope_se"))


# Read gtex heart egene list
# Only consider those that were tested in both species and are one2one orthologs
GtexHeartEgenes <- read.table("../data/Heart_Left_Ventricle.v7.egenes.txt.gz", header=T, sep='\t', stringsAsFactors = F) %>% 
  left_join(HumanLeadSnps, by=c("gene_id"="gene")) %>%
  mutate(gene_id_stable = gsub(".\\d+$","",gene_id)) %>%
  filter(gene_id_stable %in% OneToOneMap$Gene.stable.ID) %>%
  mutate(chimp_id = plyr::mapvalues(gene_id_stable, OneToOneMap$Gene.stable.ID,  OneToOneMap$Chimpanzee.gene.stable.ID, warn_missing = F)) %>%
  filter(chimp_id %in% ChimpTestedGenes)

ChimpToHuman.ID <- function(Chimp.ID){
  #function to convert chimp ensembl to human ensembl gene ids
  return(
    plyr::mapvalues(Chimp.ID, OneToOneMap$Chimpanzee.gene.stable.ID, OneToOneMap$Gene.stable.ID, warn_missing = F)
  )}
```


First question: do the ~300 chimp eGenes have more contacts in their cis-window in chimp

```{r}

Chimp_OrderedGenes <- eQTLs %>%
    mutate(HumanID = ChimpToHuman.ID(gene)) %>%
    group_by(gene) %>% 
    dplyr::slice(which.min(qvalue)) %>%
    filter(gene %in% GtexHeartEgenes$chimp_id) %>%
    left_join(GtexHeartEgenes, by=c("gene"="chimp_id")) %>%
    dplyr::select(gene, qvalue, p, qval, HumanID) %>% as.data.frame() %>%
    filter(qvalue <0.1) %>%
    mutate(ChimpRank = dense_rank(qvalue)) %>%
    mutate(ChimpRelativeRank = ChimpRank/max(ChimpRank)) %>%
    mutate(HumanRank = dense_rank(qval)) %>%
    mutate(HumanRelativeRank = HumanRank/max(HumanRank)) %>%
    mutate(RankDifference = HumanRank-ChimpRank)

# OneToOneMap %>%
#   inner_join(HumanInteractions, by=c("Gene.stable.ID"="HumanId")) %>% dim()
#   inner_join(ChimpInteractions, by=c("Chimpanzee.gene.stable.ID"="ChimpID")) %>% dim()
#   right_join(Chimp_OrderedGenes, by=c("Chimpanzee.gene.stable.ID"="gene")) %>% dim()

Chimp_OrderedGenes.WithContactInfo <- Chimp_OrderedGenes %>%
  left_join(HumanInteractions, by=c("HumanID")) %>%
  left_join(ChimpInteractions, by=c("gene"="ChimpID")) %>% 
  mutate(InteractionDifference=H.Score - C.Score)

  
ggplot(Chimp_OrderedGenes.WithContactInfo, aes(x=InteractionDifference)) +
  stat_ecdf(geom = "step") +
  xlab("Difference in contacts over chimp eGene cis-windows\n(Positive means more contact in chimp)") +
  ylab("Cumulative frequency") +
  theme_bw()
  
ggplot(Chimp_OrderedGenes.WithContactInfo, aes(x=RankDifference, y=InteractionDifference)) +
  geom_point() +
  theme_bw() +
  xlab("Rank Difference in eGene significance\nMore in human <--  --> More in chimp") +
  ylab("Differential contacts in cis window\nMore in human <--  --> More in chimp") +
  geom_smooth(method='lm',formula=y~x)

cor.test(x=Chimp_OrderedGenes.WithContactInfo$RankDifference, y=Chimp_OrderedGenes.WithContactInfo$InteractionDifference, method="spearman")

contacts.v.eGene.lm = lm(InteractionDifference ~ RankDifference, data=Chimp_OrderedGenes.WithContactInfo)
summary(contacts.v.eGene.lm)
plot(contacts.v.eGene.lm)
```

With this procedure, the correlation was weaker. The earlier observation, that chimp eGenes have more dna contacts in their cis-windows in chimp, is not robustly detected.


The perhaps more sensitive way to ask a slightly different question, is this: do chimp eQTL snps (or conversely human eQTL snps), have more contacts between TSS and SNP in its respective species?
