---
title: "Final_3_GSEA"
author: "Benjamin Fair"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

## Introduction

This markdown will be used to generate publication quality figures that I have already made at some point in my exploratory Rmarkdown analyses. This will mostly be plots that have to do with gene ontology or gene set association analysis.

## Analysis

First, load necessary libraries and read in data
```{r}
library(tidyverse)
library("clusterProfiler")
library("org.Hs.eg.db")
library(enrichplot)
library(stringr)
source("../code/CustomFunctions.R")
```

read in eGene data

```{r}
EgenesTested <- TsvToCombinedEgenes(Chimp.tsv = "../output/ChimpEgenes.eigenMT.txt.gz", Human.tsv = "../data/GTEX_v8_eGenes/Heart_Left_Ventricle.v8.egenes.txt.gz", SysToID.tsv = "../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", HumanTsvType = "GTEx")
EgenesTested.grouped <- AddGroups(EgenesTested, HumanEgeneCount=495)

Dispersion <- read.delim('../output/OverdispersionEstimatesFromChimp.txt')
Dispersion.se <- read.delim('../output/OverdispersionEstimatesFromChimp.txt.SE.tab.gz')
Dispersion.P <- read.delim('../output/OverdispersionEstimatesFromChimp.txt.Pvals.tab')
Dispersion %>%
  dplyr::select(gene,
                Chimp.Mean.Expression,
                Human.Mean.Expression,
                Chimp.Overdispersion,
                Human.Overdispersion,
                Chimp.Mean.Adjusted.Dispersion=Chimp.Residual,
                Human.Mean.Adjusted.Dispersion=Human.Residual) %>%
  left_join(Dispersion.se, by="gene") %>%
  left_join(Dispersion.P, by="gene") %>% head()
```

First define a function to do GSEA on Molecular Function (MF), Cellular Component (CC), and Biological process (BP).

```{r}
gsea.full <- function(NamedGeneVector, ...){
  
  SortedNamedGeneVector <- sort(NamedGeneVector, decreasing=T)

  BP<-gseGO(gene=SortedNamedGeneVector,
            ont = "BP",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
  BiologicalProcess<-as.data.frame(simplify(BP)) %>%
    mutate(ONTOLOGY="BP")

  MF<-gseGO(gene=SortedNamedGeneVector,
            ont = "MF",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
  MolecularFunction<-as.data.frame(simplify(MF)) %>%
    mutate(ONTOLOGY="MF")

  CC<-gseGO(gene=SortedNamedGeneVector,
            ont = "CC",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
  CellularComponent<-as.data.frame(simplify(CC)) %>%
    mutate(ONTOLOGY="CC")

  
  Combined.simplified <- combine(BiologicalProcess,MolecularFunction,CellularComponent)
  
  return(list(BP=BP, MF=MF, CC=CC, Simplified=Combined.simplified))
}
```


Perform GSEA based on difference in dispersion. What genes are more dispersed in chimp than human, vice-versa

```{r}
ChimpOverdispersionTable <- read.table("../output/OverdispersionEstimatesFromChimp.txt", header=T, sep='\t', stringsAsFactors = F)

DifferenceInDispersion <- ChimpOverdispersionTable %>%
  mutate(DispersionDiff = Chimp.Residual-Human.Residual ) %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
  dplyr::select(gene, DispersionDiff) %>%
  deframe() %>% sort(decreasing = T)

gsea.difference.in.disp<-gseGO(gene=DifferenceInDispersion,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)

# 
# Full.gsea.simplified <- gsea.full(DifferenceInDispersion)
# 
# Full.gsea.simplified$Simplified %>% dim()
# Full.gsea.simplified$Simplified %>% dplyr::select(Description)

DispersionContrastPlot <-
  # Full.gsea.simplified$Simplified %>% as.data.frame() %>%
  gsea.difference.in.disp %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  # filter(str_detect(Description, 'DNA')) %>%
  # group_by(ONTOLOGY, Polarization) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n = 3, wt = abs(enrichmentScore)) %>%
  ungroup() %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  scale_x_continuous(limits=c(-1,1), breaks=c(-1,0,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.0001, 0.1), direction=-1, option="D") +
  # scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  # xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  xlab("Enrichment") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 35, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw() +
  theme(axis.text.y = element_text(size=6)) + 
  ylab(NULL)
DispersionContrastPlot

ggsave("../figures/OriginalArt/GSEA.dispersionContrast.eps", DispersionContrastPlot, width=4, height=4)

write_delim(as.data.frame(gsea.difference.in.disp) %>% dplyr::select(-core_enrichment), path="../output/Final/TableS5.tab", delim='\t')


# Full.gsea.simplified$BP %>% as.data.frame() %>% dplyr::select(p.adjust, enrichmentScore, Description) %>%
# filter(grepl("DNA",Description))


gsea.difference.in.disp %>% as.data.frame() %>%
# Full.gsea.simplified$Simplified %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n=-4, wt=p.adjust) %>%
  top_n(n=4, wt=enrichmentScore) %>%
  ungroup() %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.0001, 0.1), direction=-1, option="D") +
  facet_grid(ONTOLOGY~., scales = "free") +
  xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 60, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw()



```

Differential GSEA based on polarized P-value

```{r}
DifferenceInDispersion <- ChimpOverdispersionTable %>%
  left_join(Dispersion.P, by="gene") %>%
  mutate(Polarized.P = -log10(P) * sign(Chimp.Residual-Human.Residual)) %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
  dplyr::select(gene, Polarized.P) %>%
  deframe() %>% sort(decreasing = T)

gsea.difference.in.disp<-gseGO(gene=DifferenceInDispersion,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)

DispersionContrastPlot <-
  # Full.gsea.simplified$Simplified %>% as.data.frame() %>%
  gsea.difference.in.disp %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  # filter(str_detect(Description, 'DNA')) %>%
  # group_by(ONTOLOGY, Polarization) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n = 5, wt = abs(enrichmentScore)) %>%
  ungroup() %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  scale_x_continuous(limits=c(-1,1), breaks=c(-1,0,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.0001, 0.1), direction=-1, option="D") +
  # scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  # xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  xlab("Enrichment") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 35, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw() +
  theme(axis.text.y = element_text(size=6)) + 
  ylab(NULL)
DispersionContrastPlot

ggsave("../figures/OriginalArt/GSEA.dispersionContrast.PolarizedP.eps", DispersionContrastPlot, width=4, height=4)
```


Ok another way to do it is as GO enrichment analysis based on significant interspecies variability changes only.

```{r}
ChimpOverdispersionTable %>%
  filter((Chimp.Residual-Human.Residual)>0 ) %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
```


Now GSEA of dispersed genes based on human.

```{r}
HumanDispersion <- ChimpOverdispersionTable %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
  dplyr::select(gene, Human.Residual) %>%
  deframe() %>% sort(decreasing = T)

gsea.ranked.disp<-gseGO(gene=HumanDispersion,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=1000000)

DispersionHumanPlot <-
  # Full.gsea.simplified$Simplified %>% as.data.frame() %>%
  gsea.ranked.disp %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize, NES) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  # filter(str_detect(Description, 'DNA')) %>%
  # group_by(ONTOLOGY, Polarization) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n = 3, wt = abs(enrichmentScore)) %>%
  
  # top_n(n = 3, wt = abs(-log10(p.adjust))) %>%
  # top_n(n = 3, wt = abs(enrichmentScore)) %>%

  ungroup() %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  scale_x_continuous(limits=c(-1,1), breaks=c(-1,0,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.00001, 0.1), direction=-1, option="D") +
  # scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  # xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  xlab("Enrichment") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 35, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw() +
  theme(axis.text.y = element_text(size=6)) + 
  ylab(NULL)
DispersionHumanPlot

ggsave("../figures/OriginalArt/GSEA.dispersionHuman.eps", DispersionHumanPlot, width=4, height=4)

write_delim(as.data.frame(gsea.ranked.disp) %>% dplyr::select(-core_enrichment), path="../output/Final/TableS4.tab", delim='\t')

```

Repeat but excluding viral challanged chimps

```{r}
ChimpOverdispersionTableNoVirus <- read.table("../output/OverdispersionEstimatesFromChimp_NoVirusChallangedIndividuals.txt", header=T, sep='\t', stringsAsFactors = F)

DifferenceInDispersionNoVirus <- ChimpOverdispersionTableNoVirus %>%
  mutate(DispersionDiff = Chimp.Residual-Human.Residual ) %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
  dplyr::select(gene, DispersionDiff) %>%
  deframe() %>% sort(decreasing = T)

gsea.difference.in.disp.NoVirus<-gseGO(gene=DifferenceInDispersionNoVirus,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
DispersionContrastPlotNoVirus <-
  # Full.gsea.simplified$Simplified %>% as.data.frame() %>%
  gsea.difference.in.disp.NoVirus %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize, NES) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  # filter(str_detect(Description, 'DNA')) %>%
  # group_by(ONTOLOGY, Polarization) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n = 3, wt = abs(enrichmentScore)) %>%
  # top_n(n = 4, wt = abs(-log10(p.adjust))) %>%
  ungroup() %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  scale_x_continuous(limits=c(-1,1), breaks=c(-1,0,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.0001, 0.1), direction=-1, option="D") +
  # scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  # xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  xlab("Enrichment") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 35, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw() +
  theme(axis.text.y = element_text(size=6)) + 
  ylab(NULL)
DispersionContrastPlotNoVirus

ggsave("../figures/OriginalArt/GSEA.dispersionContrastNoVirus.eps", DispersionContrastPlotNoVirus, width=4, height=4)

write_delim(as.data.frame(gsea.difference.in.disp.NoVirus) %>% dplyr::select(-core_enrichment), path="../output/Final/TableS6.tab", delim='\t')

HumanDispersionNoVirus <- ChimpOverdispersionTableNoVirus %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
  dplyr::select(gene, Human.Residual) %>%
  deframe() %>% sort(decreasing = T)

gsea.ranked.disp.NoVirus<-gseGO(gene=HumanDispersionNoVirus,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=1000000)

DispersionHumanPlotNoVirus <-
  # Full.gsea.simplified$Simplified %>% as.data.frame() %>%
  gsea.ranked.disp.NoVirus %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  # filter(str_detect(Description, 'DNA')) %>%
  # group_by(ONTOLOGY, Polarization) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n = 3, wt = abs(enrichmentScore)) %>%
  ungroup() %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  scale_x_continuous(limits=c(-1,1), breaks=c(-1,0,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.00001, 0.1), direction=-1, option="D") +
  # scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  # xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  xlab("Enrichment") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 35, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw() +
  theme(axis.text.y = element_text(size=6)) + 
  ylab(NULL)
DispersionHumanPlotNoVirus

```


Now, Perform differential GSEA based on difference in qvalue ranks for eGene between species. I set eval=F to this code since it takes so long and based on the fact that the difference in FDR doesn't correlate with dispersion, I think this is convoluted and a generally not good way to ask the question "what genes are more eGene-y in chimp than human?"

```{r, eval=F}

DifferenceInFDR.Rank <- EgenesTested.grouped %>%
  mutate(H.rank = dense_rank(H.FDR),
         C.rank = dense_rank(C.FDR)) %>%
  filter(C.FDR <= 0.1) %>%
  mutate(RankDiff=C.rank-H.rank) %>%
  dplyr::select(H.gene, RankDiff) %>%
  deframe() %>% sort(decreasing = T)

gsea.difference.in.fdr <-gseGO(gene=DifferenceInFDR.Rank,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=500000,
            pAdjustMethod='fdr',
            pvalueCutoff=0.1)

gsea.difference.in.fdr %>% as.data.frame() %>%
  dplyr::select(Description)

gsea.difference.in.fdr %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 60, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw()

gseaplot2(gsea.difference.in.fdr, geneSetID = c("GO:0006366","GO:0006355"), title="Chimp eGenes ordered by difference in species rank", pvalue_table = TRUE)

#Compare gene lists using fdr vs dispersion
A<- gsea.difference.in.fdr %>% as.data.frame() %>% pull(ID)

B<- gsea.difference.in.disp %>% as.data.frame() %>% pull(ID)

length(A)
length(B)
length(intersect(A,B))

```

None of the same categories.

Since the difference in FDR rank appraoch is convoluted, I will ask that same question ("What genes are more of eGenes in chimp than in human") by doing gene overlap analysis using chimp specific eGenes as foreground, and all chimp eGenes as background.


```{r}

EgenesTested.FDR10 <- AddGroups(TsvToCombinedEgenes(Chimp.tsv = "../output/ChimpEgenes.eigenMT.txt.gz", Human.tsv = "../data/GTEX_v8_eGenes/Heart_Left_Ventricle.v8.egenes.txt.gz", SysToID.tsv = "../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", HumanTsvType = "GTEx"))


EgenesTested.FDR10 %>% pull(group) %>% table()

Foreground <- EgenesTested.FDR10 %>%
  filter(group %in% c("chimp")) %>% pull(H.gene)

Background <- EgenesTested.FDR10 %>%
  filter(group %in% c("chimp", "both")) %>% pull(H.gene)

ChimpSpecificEgene.GO <- enrichGO(gene = Foreground,
                 universe = Background,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 ont = "ALL")

dotplot(ChimpSpecificEgene.GO)

ChimpEgeneGO.plot <- ChimpSpecificEgene.GO %>% as.data.frame() %>%
  separate(GeneRatio, c("FgY", "BgY"), sep="/", convert=TRUE) %>%
  separate(BgRatio, c("FgN", "BgN"), sep="/", convert=TRUE) %>%
  mutate(OddsRatio=(FgY/BgY)/(FgN/BgN)) %>%
  dplyr::filter(p.adjust < 0.1) %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  dplyr::select(ONTOLOGY, Description, p.adjust, qvalue, pvalue, OddsRatio) %>%
  ggplot(aes(x=Description, y=OddsRatio, fill=p.adjust)) +
  geom_bar(stat='identity') +
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 40, simplify = FALSE), paste, collapse="\n")) +
  scale_fill_viridis_c(trans="log10", limits=c(0.01, 1), direction=-1, option="D") +

  coord_flip() +
  geom_hline(yintercept = 1, linetype="dotted") +
  facet_grid(ONTOLOGY~., scales = "free") +
  theme_bw() +
  theme(axis.title.y=element_blank())
ggsave("../figures/OriginalArt/GO.ChimpSpecificEgenes.eps", ChimpEgeneGO.plot, width=4.5, height=4.5)

GeneSymbols <- read.delim("../data/HumanGeneIdToHGNC_Symbol.Biomart.txt.gz", stringsAsFactors = F)

p <- function(v) {
  Reduce(f=paste, x = v)
 }

ChimpSpecificEgene.GO %>% as.data.frame() %>%
  separate(GeneRatio, c("FgY", "BgY"), sep="/", convert=TRUE, remove=F) %>%
  separate(BgRatio, c("FgN", "BgN"), sep="/", convert=TRUE, remove=F) %>%
  mutate(OddsRatio=(FgY/BgY)/(FgN/BgN)) %>%
  dplyr::select(-c("FgY", "BgY", "FgN", "BgN", "Count")) %>%
  separate_rows(geneID, sep="/") %>%
  left_join(GeneSymbols, by=c("geneID"="Gene.stable.ID")) %>%
  arrange(HGNC.symbol) %>%
  group_by(ID) %>%
  dplyr::summarise(genes=paste(HGNC.symbol, collapse=", "),
                   genes_ensembl=paste(geneID, collapse=", ")) %>% 
  left_join(ChimpSpecificEgene.GO %>%
              as.data.frame() %>%
              dplyr::select(-geneID),
            by="ID") %>%
  mutate(genes=str_remove(genes, "^, ")) %>%
  dplyr::select(-genes, -genes_ensembl, everything()) %>%
  write_delim(path = "../output/Final/TableS10.tab", delim = '\t')


```

### GO overrepresentation of shared eGenes comared to chimp or human specific
```{r}

EgenesTested.grouped <- AddGroups(TsvToCombinedEgenes(Chimp.tsv = "../output/ChimpEgenes.eigenMT.txt.gz", Human.tsv = "../data/GTEX_v8_eGenes/Heart_Left_Ventricle.v8.egenes.txt.gz", SysToID.tsv = "../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", HumanTsvType = "GTEx"), HumanEgeneCount=495)

Foreground <- EgenesTested.grouped %>%
  filter(group %in% c("both")) %>% pull(H.gene)

Background <- EgenesTested.grouped %>%
  filter(group %in% c("chimp", "both", "human")) %>% pull(H.gene)

Shared.eGene.GO <- enrichGO(gene = Foreground,
                 universe = Background,
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 ont = "ALL", 
                 pAdjustMethod = "BH", 
                 pvalueCutoff=0.05)
SharedEgeneGO.plot <- Shared.eGene.GO %>% as.data.frame() %>%
  separate(GeneRatio, c("FgY", "BgY"), sep="/", convert=TRUE) %>%
  separate(BgRatio, c("FgN", "BgN"), sep="/", convert=TRUE) %>%
  mutate(OddsRatio=(FgY/BgY)/(FgN/BgN)) %>%
  dplyr::filter(p.adjust < 0.1) %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  dplyr::select(ONTOLOGY, Description, p.adjust, qvalue, pvalue, OddsRatio) %>%
  group_by(ONTOLOGY) %>%
  top_n(n = 5, wt = OddsRatio) %>%
  ungroup() %>%
  ggplot(aes(x=Description, y=OddsRatio, fill=p.adjust)) +
  geom_bar(stat='identity') +
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 40, simplify = FALSE), paste, collapse="\n")) +
  scale_fill_viridis_c(trans="log10", limits=c(0.0001, 1), direction=-1, option="D") +

  coord_flip() +
  geom_hline(yintercept = 1, linetype="dotted") +
  facet_grid(ONTOLOGY~., scales = "free") +
  theme_bw() +
  theme(axis.title.y=element_blank())

SharedEgeneGO.plot
ggsave("../figures/OriginalArt/GO.SharedEgenes.eps", SharedEgeneGO.plot, width=4.5, height=4.5)

Shared.eGene.GO %>% as.data.frame() %>%
  separate(GeneRatio, c("FgY", "BgY"), sep="/", convert=TRUE, remove=F) %>%
  separate(BgRatio, c("FgN", "BgN"), sep="/", convert=TRUE, remove=F) %>%
  mutate(OddsRatio=(FgY/BgY)/(FgN/BgN)) %>%
  dplyr::select(-c("FgY", "BgY", "FgN", "BgN", "Count")) %>%
  separate_rows(geneID, sep="/") %>%
  left_join(GeneSymbols, by=c("geneID"="Gene.stable.ID")) %>%
  arrange(HGNC.symbol) %>%
  group_by(ID) %>%
  dplyr::summarise(genes=paste(HGNC.symbol, collapse=", "),
                   genes_ensembl=paste(geneID, collapse=", ")) %>% 
  left_join(Shared.eGene.GO %>%
              as.data.frame() %>%
              dplyr::select(-geneID),
            by="ID") %>%
  mutate(genes=str_remove(genes, "^, ")) %>%
  dplyr::select(-genes, -genes_ensembl, everything()) %>%
  write_delim(path = "../output/Final/TableS9.tab", delim = '\t')

```

## Conclusions


