---
title: "Final_4_DGN"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

## Introduction

This markdown will be used to generate publication quality figures that I have already made at some point in my exploratory Rmarkdown analyses. This will mostly be plots that have to do with gene ontology or gene set association analysis.

## Analysis

First, load necessary libraries and read in data
```{r}
library(tidyverse)
library("clusterProfiler")
library("org.Hs.eg.db")
library(enrichplot)
library(stringr)
source("../code/CustomFunctions.R")
library(DOSE)
```

read in eGene data

```{r}
EgenesTested <- TsvToCombinedEgenes(Chimp.tsv = "../output/ChimpEgenes.eigenMT.txt.gz", Human.tsv = "../data/GTEX_v8_eGenes/Heart_Left_Ventricle.v8.egenes.txt.gz", SysToID.tsv = "../data/Biomart_export.Hsap.Ptro.orthologs.txt.gz", HumanTsvType = "GTEx")
EgenesTested.grouped <- AddGroups(EgenesTested, HumanEgeneCount=495)
```

First define a function to do GSEA on Molecular Function (MF), Cellular Component (CC), and Biological process (BP).

```{r, eval=F}
gsea.full <- function(NamedGeneVector, ...){
  
  SortedNamedGeneVector <- sort(NamedGeneVector, decreasing=T)

  BP<-gseGO(gene=SortedNamedGeneVector,
            ont = "BP",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
  BiologicalProcess<-as.data.frame(simplify(BP)) %>%
    mutate(ONTOLOGY="BP")

  MF<-gseGO(gene=SortedNamedGeneVector,
            ont = "MF",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
  MolecularFunction<-as.data.frame(simplify(MF)) %>%
    mutate(ONTOLOGY="MF")

  CC<-gseGO(gene=SortedNamedGeneVector,
            ont = "CC",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)
  CellularComponent<-as.data.frame(simplify(CC)) %>%
    mutate(ONTOLOGY="CC")

  
  Combined.simplified <- combine(BiologicalProcess,MolecularFunction,CellularComponent)
  
  return(list(BP=BP, MF=MF, CC=CC, Simplified=Combined.simplified))
}
```


Perform GSEA based on difference in dispersion. What genes are more dispersed in chimp than human, vice-versa

```{r}
ChimpOverdispersionTable <- read.table("../output/OverdispersionEstimatesFromChimp_NoVirusChallangedIndividuals.txt", header=T, sep='\t', stringsAsFactors = F)

# ChimpOverdispersionTable <- read.table("../output/OverdispersionEstimatesFromChimp.txt", header=T, sep='\t', stringsAsFactors = F)

DifferenceInDispersion <- ChimpOverdispersionTable %>%
  mutate(DispersionDiff = Chimp.Residual-Human.Residual ) %>%
  filter(gene %in% EgenesTested.grouped$H.gene) %>%
  dplyr::select(gene, DispersionDiff) %>%
  deframe() %>% sort(decreasing = T)

gsea.difference.in.disp<-gseDGN(gene=DifferenceInDispersion,
            ont = "ALL",
            OrgDb=org.Hs.eg.db,
            keyType='ENSEMBL',
            nPerm=100000)


Full.gsea.simplified <- gsea.full(DifferenceInDispersion)

Full.gsea.simplified$Simplified %>% dim()
Full.gsea.simplified$Simplified %>% dplyr::select(Description)

DispersionContrastPlot <-
  # Full.gsea.simplified$Simplified %>% as.data.frame() %>%
  gsea.difference.in.disp %>% as.data.frame() %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  # filter(str_detect(Description, 'DNA')) %>%
  # group_by(ONTOLOGY, Polarization) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n = 3, wt = abs(enrichmentScore)) %>%
  ungroup() %>%
  mutate(ONTOLOGY=recode(ONTOLOGY, `BP` = "Biological Process", `CC` = "Cellular Component", `MF`="Molecular Function")) %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  scale_x_continuous(limits=c(-1,1), breaks=c(-1,0,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_viridis_c(trans="log10", limits=c(0.0001, 0.1), direction=-1, option="D") +
  # scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  # xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  xlab("Enrichment") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 35, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw() +
  theme(axis.text.y = element_text(size=6)) + 
  ylab(NULL)
DispersionContrastPlot

ggsave("../figures/OriginalArt/GSEA.dispersionContrast.eps", DispersionContrastPlot, width=4, height=4)

Full.gsea.simplified$BP %>% as.data.frame() %>% dplyr::select(p.adjust, enrichmentScore, Description) %>%
filter(grepl("DNA",Description))


gsea.difference.in.disp %>% as.data.frame() %>%
# Full.gsea.simplified$Simplified %>%
  dplyr::select(Description, ONTOLOGY, p.adjust,enrichmentScore, setSize) %>%
  mutate(Polarization=sign(enrichmentScore)) %>%
  group_by(ONTOLOGY, Polarization) %>%
  top_n(n=-4, wt=p.adjust) %>%
  top_n(n=4, wt=enrichmentScore) %>%
  ungroup() %>%
  ggplot(aes(x=enrichmentScore, y=Description, color=p.adjust, size=setSize)) +
  geom_point() +
  xlim(c(-1,1)) +
  facet_grid(ONTOLOGY~., scales = "free") +
  scale_colour_gradient(low="red", high="black") +
  facet_grid(ONTOLOGY~., scales = "free") +
  xlab("Enrichment\nOverdispersedInHuman<-->OverdispersedInChimp") +
  scale_y_discrete(labels = function(x) lapply(strwrap(x, width = 60, simplify = FALSE), paste, collapse="\n")) +
  labs(color = "Adjusted P-value") +
  geom_vline(xintercept = 0, linetype="dotted") +
  theme_bw()

```